---
title: "Lab 06 - Functions and iterations"
author: "Muhammad Sami Munawar"
date: "`r Sys.Date()`"
output: html_document
---


### Packages and Data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(ggplot2)
```

```{r diamon_rename}
diamonds <- diamonds %>% rename(
    depth_pct = depth,
    length_mm = x,
    width_mm = y,
    depth_mm = z,
    table_pct = table
  )

diamonds

```


## Exercise 1

```{r}
# Do some preliminary investigations & cleaning

# diamonds %>% count(cut) 

diamonds <- diamonds %>% filter(length_mm != 0, width_mm != 0, depth_mm != 0, na.rm = TRUE) 

summary(diamonds)

ggplot(diamonds, aes(x = length_mm, y = depth_pct)) + geom_point() # 
ggplot(diamonds, aes(x = width_mm, y = depth_pct)) + geom_point()
ggplot(diamonds, aes(x = depth_mm, y = depth_pct)) + geom_point()
```


## Exercise 2

```{r}

diamonds <- diamonds %>%
  mutate(table_mm = table_pct * width_mm / 100)

# Calculate the table width in millimetres
diamonds <- diamonds %>%
  mutate(table_mm = table_pct * width_mm / 100)

# Calculate the arithmetic mean of the table lengths for all diamonds
overall_mean <- mean(diamonds$table_mm)

# Calculate the arithmetic mean of the table lengths within each clarity category
clarity_means <- diamonds %>%
  group_by(clarity) %>%
  summarise(mean_table_mm = mean(table_mm, na.rm = TRUE))


clarity_means



```

_Add your comment here_

## Exercise 3

_Do derivation on paper, unless you know how to write LaTeX code_

Given the geometric mean formula:

\[\bar{x}_g = \sqrt[n]{\prod_{i=1}^{n} x_i}\]

We want to show that this is equal to:

\[\bar{x}_g = \exp\left(\frac{1}{n} \sum_{i=1}^{n} \ln x_i\right)\]

Starting with the right-hand side:

\begin{align*}
\exp\left(\frac{1}{n} \sum_{i=1}^{n} \ln x_i\right) &= \exp\left(\ln x_1^{1/n} + \ln x_2^{1/n} + \ldots + \ln x_n^{1/n}\right) \\
&= \exp\left(\ln \left(x_1^{1/n} x_2^{1/n} \ldots x_n^{1/n}\right)\right) \\
&= \exp\left(\ln \left(\sqrt[n]{x_1} \sqrt[n]{x_2} \ldots \sqrt[n]{x_n}\right)\right) \\
&= \exp\left(\ln \left(\sqrt[n]{x_1 x_2 \ldots x_n}\right)\right) \\
&= \sqrt[n]{x_1 x_2 \ldots x_n} \\
&= \bar{x}_g
\end{align*}


## Exercise 4

```{r}
# Write gmean() command


gmean <- function(x) {
  # Ensure that x is a numeric vector
  if (!is.numeric(x)) {
    warning("Input must be a numeric vector.")
    return("NaN")
  }
  
  # Calculate the geometric mean
  n <- length(x)
  gm <- exp(sum(log(x)) / n)
  
  return(gm)
}





```

## Exercise 5

```{r, error = TRUE}

test_data <- list(
  test1 = c(2.1, 3.8, 4.2),
  test2 = c(1, 10, 100, 1000),
  test3 = c(0, 1, 4),
  test4 = c(0.38,  0.94, -1.56),
  test5 = c(TRUE, TRUE),
  test6 = c("6", "7", "8")
)


# Create for loop to test gmean() on the above examples

for (i in seq_along(test_data)) {
  cat("gmean of test", i, ": ")
  print(gmean(test_data[[i]]))
}

```


## Exercise 6

```{r}
# Copy gmean() from Ex4 and edit it to check the input



gmean <- function(x) {
  # Ensure that x is a numeric vector
  if (!is.numeric(x)) {
    warning("Input must be a numeric vector.")
    return(NaN)
  }
  
  # Ensure that all elements of x are positive
  if (any(x <= 0)) {
    warning("All elements of input must be positive to compute geometric mean.")
    return(NaN)
  }
  
  # Calculate the geometric mean
  n <- length(x)
  gm <- exp(sum(log(x)) / n)
  
  return(gm)
}


# Copy your code from Ex 5 to test your new command

for (i in seq_along(test_data)) {
  cat("gmean of test", i, ": ")
  print(gmean(test_data[[i]]))
}



```


## Exercise 7

```{r}
# Overall table_mm geometic mean?

summary_stats <- diamonds %>%
  group_by(clarity) %>%
  summarize(
    mean_table_mm = mean(table_mm, na.rm = TRUE),
    median_table_mm = median(table_mm, na.rm = TRUE),
    geom_mean_table_mm = gmean(table_mm) # This uses the gmean function we defined earlier
  )

summary_stats



# table_mm geometic mean within each clarity category






```

_Add your comment here_


## Exercise 8 (optional)

```{r}
# Create gmean2() that computes the geometric mean 
#   that uses the original definition

gmean2 <- function(x) {
  
  if (!is.numeric(x)) {
    if (all(sapply(x, is.character))) {
      x <- as.numeric(x)  # attempt to convert characters to numeric
    }
  }
  
  # Check for non-numeric values
  
    if (any(is.na(x))) {
      stop("Input must be a numeric vector.")
    }
  
  
  # Check for non-positive values
  if (any(x <= 0, na.rm = TRUE)) {
    warning("Values must be strictly positive for geometric mean calculation.")
    return(NaN)
  }
  
  # Calculate the geometric mean
  exp(mean(log(x), na.rm = TRUE))
}

# Apply the gmean2 function on the test data
for (i in seq_along(test_data)) {
  cat("gmean2 of test", i, ": ")
  print(gmean2(test_data[[i]]))
}




```


## Exercise 9 (optional)

```{r}
# Create a function that computes the harmonic mean


hmean <- function(x) {
  if (any(x <= 0, na.rm = TRUE)) {
    warning("All values must be positive for harmonic mean calculation.")
    return(NaN)
  }
  if (!is.numeric(x)) {
    warning("Input must be a numeric vector.")
    return(NaN)
  }
  n <- sum(!is.na(x)) # Count of non-NA values
  hm <- n / sum(1 / x, na.rm = TRUE)
  return(hm)
}


for (i in 1:length(test_data)) {
  cat("Harmonic mean of test", i, ":", hmean(test_data[[i]]), "\n")
}


```
